---
title: "Data Cleaning Workflow"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{vignette}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```


```{r setup}
library(whiskerTrimmer)
```


# Introduction

whiskerTrimmer provides a set of functions designed to clean and preprocess experimental mouse data. Common issues addressed include:

* Fixing incorrectly recorded mouse gender IDs.
* Cleaning columns that contain mixed data types (numeric and non-numeric).
* Detecting and handling outliers using various statistical methods.
* Identifying potential misspellings in categorical data.
* Dropping rows with large, implausible weight reductions across measurement times.


In this vignette, we’ll demonstrate how to use these functions in a typical data cleaning workflow.

# Prerequisites
Before starting, ensure you have:

* Installed and loaded whiskerTrimmer

```{r}
# install.packages("devtools") if needed
devtools::load_all()
```
* A dataset that you want to clean, structured with a column for mouse IDs and various measurement columns.

For demonstration, let’s create a small, hypothetical dataset. In your real workflow, you might load your data using `read_excel()` or `read.csv()`.

```{r}

mouse_data <- data.frame(
  ID = c("Mouse_F_01", "Mouse_F_02", "Mouse_M_03", "Mouse_F_04"),
  Weight_Day1 = c("20g", "19", "twenty", "21"),
  Weight_Day2 = c("19.5g", "20", "twenty one", "16"),
  Diet = c("Stadard", "Standard", "Stanrd", "Standard") # misspellings of "Standard"
)
mouse_data
```

# Step-by-Step Cleaning

1. Fixing Gender Patterns

`fix_gender_pattern()` corrects single mis-typed genders in mouse IDs. It expects a pattern like Prefix_Gender_Number (e.g., Mouse_F_01), and if one entry is out of place, it fixes it to match the neighboring entries.

```{r}
# Run fix_gender_pattern()
cleaned_data <- fix_gender_pattern(data = mouse_data, column_name = "ID")
cleaned_data
```

2. Cleaning Mixed Columns

`clean_mixed_columns()` removes non-numeric characters from columns that should be numeric. Here, Weight_Day1 and Weight_Day2 contain textual elements like "g" or words.

```{r}
cleaned_data <- clean_mixed_columns(cleaned_data, Weight_Day1, Weight_Day2)
cleaned_data
```

3. Detecting Outliers

`detect_outliers()` identifies outliers using either a mean-based (Z-score) or median-based (IQR) method. Here, we use the Z-score method `(method = "mean")` with a threshold of 2.2. Set `drop_outliers = TRUE` to replace outliers with `NA`.

```{r}
outlier_detection <- detect_outliers(cleaned_data, Weight_Day1, Weight_Day2,
                                     method = "mean", threshold = 2.2, drop_outliers = TRUE)
outlier_detection$Results   # Outlier details
cleaned_data <- outlier_detection$Cleaned_Data
cleaned_data
```

4. Checking for Misspellings

`check_misspellings()` helps identify values that may be close matches to a more common value. Suppose `Diet` should always be "Standard" but contains variations like "Stadard" or "Stanrd":

```{r}
misspellings <- check_misspellings(cleaned_data, "Diet", frequency_threshold = 3, similarity_threshold = 0.8)
misspellings
```


5. Dropping Rows with Large Weight Reductions

`drop_large_weight_reductions()` removes rows where weights drop too sharply between measurements. For instance, a >20% drop between days might be unrealistic.

```{r}
weight_adjustment <- drop_large_weight_reductions(cleaned_data, Weight_Day1, Weight_Day2, threshold = 20)
weight_adjustment$Dropped_Summary
cleaned_data <- weight_adjustment$Cleaned_Data
cleaned_data
```

6. Plot Mouse Body Weight Over Time by Treatment Group

`plot_body_weight` function is designed to generate an insightful visualization of mouse body weight trends over time, segmented by treatment groups and sex. This function facilitates the analysis of experimental data by transforming the input dataset into a format suitable for visualization and plotting using `ggplot2`.

Function Details

The function performs the following steps:

* Reshape the Data: Transforms the dataset from wide to long format using the pivot_longer function to make it compatible with ggplot2.

* The weight columns are combined into a single weight column.

* The date columns are combined into a single date column.

* Filter Matched Measurements: Ensures that weights and dates with matching suffixes are paired correctly.

* Generate the Plot: Creates a line plot of body weight over time for each mouse, grouped by treatment and faceted by sex.

Example Usage:

```{r}
library(tidyr)
library(ggplot2)
sample_data <- data.frame(
  ID = c(1, 2, 3, 4),
  sex = c("M", "F", "M", "F"),
  treatment = c("A", "B", "A", "B"),
  weight_1 = c(25, 22, 30, 28),
  weight_2 = c(26, 23, 31, 29),
  date_1 = as.Date(c("2024-01-01", "2024-01-01", "2024-01-01", "2024-01-01")),
  date_2 = as.Date(c("2024-02-01", "2024-02-01", "2024-02-01", "2024-02-01"))
)

plot <- plot_body_weight(sample_data)
```

7. Generate a Summary Table of Average Body Weights

The `generate_summary_table function` is a utility for creating summary tables that display the average body weights of mice grouped by treatment and sex. This allows researchers to easily identify trends and differences in experimental data.

The function performs the following steps:

*Group the Data: Groups the dataset by treatment and sex.

*Calculate Averages: Computes the mean weight for each group at different time points, ignoring missing values.

*Rename Columns: Renames the columns for clarity in the output table.

Example Usage:

```{r}
sample_data <- data.frame(
  treatment = c("A", "B", "A", "B"),
  sex = c("M", "F", "M", "F"),
  weight_1 = c(25, 22, 30, 28),
  weight_2 = c(26, 23, 31, 29),
  weight_3 = c(27, 24, 32, 30)
)

summary_table <- generate_summary_table(sample_data)

```

# Putting it all Together

```{r}
library(whiskerTrimmer)

# Load your dataset
mouse_data <- readxl::read_excel("../inst/exdata/mousedata.xlsx", sheet = "Birth")
mouse_data_bw <- readxl::read_excel("../inst/exdata/mousedata.xlsx", sheet = "Body Weight")

# Step 1: Fix gender patterns
mouse_data <- fix_gender_pattern(mouse_data, "ID")

# Step 2: Clean numeric columns
mouse_data_bw <- clean_mixed_columns(mouse_data_bw, "Body Weight 1", "Body Weight 2", "Body Weight 3")

# Step 3: Detect and handle outliers
outlier_detection <- detect_outliers(mouse_data_bw, "Body Weight 1", "Body Weight 2", "Body Weight 3",
                                     method = "mean", threshold = 2.2, drop_outliers = TRUE)
mouse_data_bw <- outlier_detection$Cleaned_Data

# Step 4: Identify potential misspellings
misspellings <- check_misspellings(mouse_data, "Treatment", frequency_threshold = 3, similarity_threshold = 0.8)
print(misspellings)

# Step 5: Drop rows with large weight reductions
weight_adjustment <- drop_large_weight_reductions(mouse_data_bw, "Body Weight 1", "Body Weight 2", "Body Weight 3", threshold = 20)
mouse_data_bw <- weight_adjustment$Cleaned_Data

# Now `mouse_data` is cleaner and more consistent.
```

