---
title: "nfl_data_cleaning"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{nfl_data_cleaning}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)

library(dplyr)
library(ggplot2)
library(patchwork)
library(rlang)
```

```{r setup}
library(whiskerTrimmer)
```

#Introduction
This vignette demonstrates functions designed to work with NFL play-by-play or similar datasets. The functions illustrated here:

1. `na_summary()`: Summarizes missing values (NAs) in each column.
2. `visualize_outliers()`: Creates pairwise plots of numeric columns and highlights potential outliers.
3. `true_yardline()`: Calculates a "true yardline" to account for how the NFL designates field position.


# Example Data
For demonstration, let’s create a small, synthetic NFL-like dataset. In reality, you would load your dataset from a CSV, Excel file, or API:

```{r}
data <- data.frame(
  possessionTeam = c("NE", "NE", "DAL", "DAL", "NE"),
  yardlineSide = c("NE", "DAL", "DAL", "NE", "NE"),
  yardlineNumber = c(30, 45, 10, 48, NA),
  yardsToGo = c(10, 8, 7, NA, 10),
  PassLength = c(NA, 15, 5, 25, 0),
  YardsAfterCatch = c(5, NA, 0, 10, 2)
)

data
```
# Summarizing Missing Values

`na_summary()` provides a quick overview of how missing values are distributed across the dataset. It reports:

* Number of NAs in each column.
* Percentage of NAs in that column.
* How many NAs are "row-unique" (the only NA in that row).
* The mean and standard deviation for numeric columns (if applicable).

```{r}
na_info <- na_summary(data)
na_info
```

# Visualizing Outliers

`visualize_outliers()` helps you understand the distribution of numeric columns and identify potential outliers. It uses a Z-score method by default, marking points that exceed the threshold (`default z_threshold = 3`) as outliers.

Pairwise plots are created so that:

* Diagonal panels show histograms of individual columns.
* Off-diagonal panels show scatterplots of pairs of columns, with potential outliers highlighted in red.

```{r}
plot_outliers <- visualize_outliers(data, yardsToGo, PassLength, YardsAfterCatch)
plot_outliers
```

# Calculating the True Yardline


In NFL data, yardline numbering changes at the 50-yard line. The function `true_yardline()` computes a continuous measure of field position from the offense’s perspective:

* If the offense is on its own side, the yardline is the given number (e.g., their own 30-yard line is 30).
* If the offense has crossed into the opponent’s side of the field, the yardline is computed as `100 - yardlineNumber`.


```{r}
true_position <- true_yardline(data, "possessionTeam", "yardlineSide", "yardlineNumber")
true_position
```



